Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2018-12-14 Manuscript/manuscript.Rmd
MD review 12-14-2018

The mechanisms underlying the formation of TADs are a complex and active area of research. Recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangements of higher-order chromatin architecture [@phillips2013chromatin]. Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long-range chromatin contacts [@zuin2014cohesin]. Other regulatory elements like particular histone modifications such as H3k36 and H3K27 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions [@rao20143d]. These have been found to be associated with more open chromatin which alter the accessibility of genes for transcription. The distinct patterns of different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the development of TAD boundaries, allowing us to better understand what leads to their formation.

A study published by Mourad et al. used multiple logistic regression with LASSO regularization to predict TAD boundaries [@mourad2016computational]. They considered binning the genome at both 50 base and 1kb intervals and flanking TAD boundaries by 1kb and 20kb regions. The outcome Y was determined by whether or not a genomic bin overlapped with a flanked boundary. The feature space was obtained by considering the percent overlap between a bin and coordinates defining ChIP-seq peak data. Models were then evaluated using AUROC. However, a pervasive issue here becomes the introduction of highly imbalanced classes. Due to the sparcity of TADs throughout the genome, most bins will not overlap with a flanked TAD boundary. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data. The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a genomic bin overlapping with a TAD boundary region--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance.

@@

#METHODS

##DOMAIN DATA

Replicate Hi-C data for the GM12878 cell line were downloaded from GEO GSE63525 (Experiments HIC001 to HIC029) [@rao20143d]. The chromosomal coordinates of topologically associating domains throughout the genome were obtained using the arrowhead algorithm from the Juicer toolbox provided by the Aiden Laboratory (https://github.com/aidenlab/juicer) [@durand2016juicebox]. Using the juicer toolbox, we were able to call TADs on 4 different contact matrix resolutions, which included 5 kb, 25 kb, 50 kb, and 100 kb. Identified TAD boundaries were represented by their location on the linear genome (hg19 human genome assembly), including start and end coordinates for chromosomes 1 through 22 (Figure 18 A). The start and end coordinates of all hierarchical TADs were concatenated, sorted, and unique boundary points representing the borders demarcating TADs were obtained (Figure 18 B).

##MODEL CONSTRUCTION

To establish the data for modeling, the genome was binned into equally sized intervals (Figure 18 C). In order for a more systematic approach, the binning process was performed at the same resolution of the contact matrix that the domain data was obtained from. For example, for the domain data obtained from a 5kb resolution contact matrix, the genome was subsequently binned into 5kb intervals. The same processes was repeated for each of the other three resolutions. Again, operating systematically, we flanked the boundary coordinates on either side by a unit corresponding to the resolution of the particular domain data. For example, for the domain data obtained from a 5kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 5kb for a total width around each boundary point of 10kb. The same process was repeated for the other three resolutions. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

##GENOMIC ANNOTATIONS

@@

###FEATURE ENGINEERING USING GENOMIC ANNOTATIONS

@@

##ENSEMBLE FRAMEWORK

@@

###TRAINING & TESTING SETS

@@

###VARIABLE REDUCTION

@@

###RE-SAMPLING TECHNIQUES

@@

###CLASSIFICATION

@@

Once a system was defined, a random forest algorithm was performed. Letting |X| denote the cardinality of each set, then we evaluated |R| x |P| x |V| x |C| different ensemble systems in this paper.

###MODEL EVALUATION & PREDICTION

@@

$$
\begin{equation}
\begin{aligned}
Accuracy &= \dfrac{TP + TN}{TP + TN + FP + FN} \\
F1 - Score &= \dfrac{2TP}{2TP + FP + FN} \\
MCC &= \dfrac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}
\end{aligned}
\end{equation}
$$
Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set.  We also considered the AUC for each model by computing the average trapezoidal approximations for the curve created by the true positive rates (TPR) and false positive rates (FPR).

@@

#RESULTS

##ENSEMBLE COMPARISONS

@@

###CLASS DISTRIBUTIONS ACROSS RESOLUTIONS

@@

###CLASS IMBALANCE HINDERS PERFORMANCE

@@

###RANDOM UNDER-SAMPLING IMPROVES MODEL PERFORMANCE

@@

###DISTANCE TYPE PREDICTORS OUT-PERFORM OTHER PREDICTOR TYPES

@@

###PERFORMANCES DEGRADE AS RESOLUTION INCREASES

>>>>

The mechanisms underlying the formation of TADs are a complex and active area of research. Recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangements of higher-order chromatin architecture [@phillips2013chromatin]. Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long-range chromatin contacts [@zuin2014cohesin]. Other regulatory elements like particular histone modifications such as H3K36 and H3K27 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions [@rao20143d]. These have been found to be associated with more open chromatin which alter the accessibility of genes for transcription. The distinct patterns of different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the development of TAD boundaries, allowing us to better understand what leads to their formation.

A study published by Mourad et al. used multiple logistic regression with LASSO regularization to predict TAD boundaries [@mourad2016computational]. They considered binning the genome at both 50 base and 1kb intervals and flanking TAD boundaries by 1kb and 20kb regions. The outcome Y was determined by whether or not a genomic bin overlapped with a flanked boundary. The feature space was obtained by considering the percent overlap between a bin and coordinates defining ChIP-seq peak data. Models were then evaluated using AUROC. However, a pervasive issue here becomes the introduction of highly imbalanced classes. Due to the sparcity of TADs throughout the genome, most bins will not overlap with a flanked TAD boundary. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data. The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a genomic bin overlapping with a TAD boundary region--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance. [??? Make paragraphs contain one meaning. Split this paragraph into two, one describes the class imbalance problem, another is Mourad paper and how they didn't address it.]

@@

# METHODS

## DOMAIN DATA

Replicate Hi-C data for the GM12878 cell line were downloaded from GEO GSE63525 (Experiments HIC001 to HIC029) [@rao20143d]. The chromosomal coordinates of topologically associating domains throughout the genome were obtained using the arrowhead algorithm from the Juicer toolbox provided by the Aiden Laboratory (https://github.com/aidenlab/juicer) [@durand2016juicebox]. Using the juicer toolbox, we were able to call TADs [??? You called TADs yourself, correct?] on 4 different contact matrix resolutions, which included 5 kb, 25 kb, 50 kb, and 100 kb. Identified TAD boundaries were represented by their location on the linear genome (hg19 human genome assembly), including start and end coordinates for chromosomes 1 through 22 (Figure 18 A). The start and end coordinates of all hierarchical TADs were concatenated, sorted, and unique boundary points representing the borders demarcating TADs were obtained (Figure 18 B).

## MODEL CONSTRUCTION

To establish the data for modeling, the genome was binned into equally sized intervals defined by the data resolution (Figure 18 C). For example, for a 5kb resolution contact matrix, the genome was subsequently binned into 5kb intervals using the coordinates of a 5kb resolution contact matrix. The same processes was repeated for each of the other three resolutions. To accommodate for the uncertainty in exact boundary location, the boundary coordinates were flanked on either side by a unit corresponding to the resolution of the corresponding contact matrix. For example, for the domain data obtained from a 5kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 5kb for a total width around each boundary point of 10kb. The same process was repeated for the other three resolutions. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

## GENOMIC ANNOTATIONS

@@

### FEATURE ENGINEERING USING GENOMIC ANNOTATIONS

@@

## ENSEMBLE FRAMEWORK

@@

### TRAINING & TESTING SETS

@@

### VARIABLE REDUCTION [??? To be modified for using LASSO only]

@@

### RE-SAMPLING TECHNIQUES

@@

### CLASSIFICATION

@@

Once a system was defined, a random forest algorithm was performed. Letting |X| denote the cardinality of each set, then we evaluated |R| x |P| x |V| x |C| different ensemble systems.

### MODEL EVALUATION & PREDICTION

@@

$$Accuracy = \frac{TP + TN}{TP + TN + FP + FN}$$
$$F1 - Score = \frac{2TP}{2TP + FP + FN}$$
$$MCC = \frac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}$$

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set.  We also considered the AUC for each model by computing the average trapezoidal approximations for the curve created by the true positive rates (TPR) and false positive rates (FPR).

@@

# RESULTS

## ENSEMBLE COMPARISONS

@@

### CLASS DISTRIBUTIONS ACROSS RESOLUTIONS

@@

### CLASS IMBALANCE HINDERS PERFORMANCE

@@

### RANDOM UNDER-SAMPLING IMPROVES MODEL PERFORMANCE

@@

### DISTANCE TYPE PREDICTORS OUT-PERFORM OTHER PREDICTOR TYPES

@@

### PERFORMANCES DEGRADE AS RESOLUTION INCREASES [??? Terminology: 5kb - high resolution. 100kb - low resolution. Adjust the text]
